Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2020-04-23 Nicolas.tex
Corrections after everyone's questions. Greg corr and comm tbd

\author{N. Nicolas
\thanks{n.nicolas@ipnl.in2p3.fr, equal contribution}
\inst{1}
    \and M. Rigault
    \thanks{m.rigault@ipnl.in2p3.fr, equal contribution}
    \inst{2}\and\\
    Y. Copin\inst{1}   
    \and R. Graziani\inst{2}
    %HERE ALPHABETIC ORDER
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and J. Nordin
    \and Y.-L. Kim\inst{1}
    \and Saul Perlmutter\inst{3}
    \and G. Smadja\inst{1}   
}

\institute{Université de Lyon, F-69622, Lyon, France; Université de Lyon
    1, Villeurbanne; CNRS/IN2P3, Institut de Physique des Deux Infinis, Lyon
    \and 
    Université Clermont Auvergne, CNRS/IN2P3, Laboratoire de
    Physique de Clermont, F-63000 Clermont-Ferrand, France.
    \and
    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720 
}

@@

In this paper, we take a step aside to probe the validity of our modeling of the
SN population, which we claim to be constituted of two age-populations
\citep{rigault2013,rigault2015,rigault2018}: one old and one younger, the former
having on average lower lightcurve stretches and being brighter after
standardization. We use the correlation between the SN age, as probed by the LsSFR, and the SN stretch to model
the expected evolution of the underlying SN stretch distribution as a function
of redshift. This modeling relies on three assumptions: (1) there are two
distinct populations of SNe~Ia; (2) the relative fraction of each of these
populations as a function of redshift follows the model presented in
\cite{rigault2018} and (3) the underlying distribution of stretch for each age
sample is constant. This paper aims at testing this specific model with datasets from the literature. We highlight

@@

SNLS typically acquires SNe~Ia in the redshift range $0.4<z<0.8$; at these
redshifts, the rest-frame Bessel $B$ band roughly corresponds to the SNLS $i$
filter, that has a $5\sigma$ depth of
24.8~mag\footnote{\href{https://www.cfht.hawaii.edu/Science/CFHTLS/cfhtlsfinalreleaseexecsummary.html}{CFHT
final release website.}}. This converts to a $z_{lim}=0.60$, in agreement with
\cite{neill2006}, \cite{perrett2010} and \cite{bazin2011}.  Fig.~14 of
\cite{perrett2010} suggests a lower limit of $z_{lim}=0.55$; both limits will be
considered, as discussed below.

@@

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008,sako2008}, which would lead to a $z_{lim}=0.24$. However, the
SDSS surveys were more sensitive to limited spectroscopic resources;
\cite{kessler2009} pointed out that during the first year of SDSS, SNe~Ia with
$r-mag<20.5$ were favored for spectroscopic follow-up, corresponding to a
redshift cut at $0.15$. For the rest of the SDSS survey, additional
spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a \mr{reasonable} completeness up to
$z_{lim}=0.2$. Following these analyses, we will use $z_{lim}=0.2$ as the
baseline SDSS redshift limit.

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift distribution of these three surveys is shown in Fig.~\ref{fig:cuts}.
As expected, the selected redshift limits roughly correspond to the peak of
these histograms.

@@

We also performed tests allowing
the high-stretch mode of the old population to differ from the young population mode, hence adding two degrees of freedom. The corresponding fit is not
significantly better, with a $\Delta$AIC of $-0.4$. 
this strengthens the fact
that the young and old populations indeed appear to share the same underlying
high-stretch mode. 
\mr{Furthermore, one might wonder whether a low-stretch mode might also exist in the young-population, see Fig.~\ref{fig:stretchlssfr}. We tested that by allowing this population to also be bi-modal, finding the amplitude of this low-stretch mode to be compatible with~0 in this young population ($<2\%$). More generally, this rises the question of inaccurate tracing of age by a given environmental tracer (here the LsSFR). A dedicated analysis be presented in Briday et al. in prep.}

@@

\subsection{Issue of with the per Sample Asymmetric modeling}
\label{sec:bbc}

Commonly used Malmquist bias correction techniques, such as the BBC-formalism, assume per Sample Asymmetric Gaussian functions for modeling the underlying stretch and color distributions.
Yet, as shown in Section~\ref{sec:results}, such a sample-based distribution is excluded as being as good as our drifting model.
Then, unlike what \citet[][section~2]{scolnic2016} and
\citet[][section~5.4]{scolnic2018a} suggested, i.e. that
traditional surveys span limited redshift ranges and therefore that the per-sample approach accounts for implicit redshift drifts, might not be correct. We stress here that, as measurements of modern surveys try to cover increasingly larger redshift ranges in order to reduce calibration systematic uncertainties, this sample-based approach  is becoming less valid, notably for PS1, DES and, soon, LSST.

We illustrate in Fig.~\ref{fig:bbc_pdf_ps1} the prediction difference in the
underlying stretch distribution between the per-sample asymmetric modeling and our Base drifting
model for the PS1 sample. Our model is bimodal and the relative amplitude of
each mode depends on the redshift-dependent fraction of old and young SNe~Ia in
the sample: the higher the fraction of old SNe~Ia (at lower redshift), the
higher the amplitude of the old-specific low-stretch mode. This redshift
dependency is shown as blue to red underlying distributions in
Fig.~\ref{fig:bbc_pdf_ps1} for redshift ranges covered by PS1. The observed
$x_1$ histogram follows our modeling defined as the sum of individual underlying SN-redshift distributions. As expected, the two modeling approaches differ mostly in the negative part of the SN stretch distribution. The asymmetric Gaussian distribution goes through the middle of the
bimodal distribution, over-estimating the number of SNe~Ia at $x_1\sim-0.7$ and
under-estimating it at $x_1\sim-1.7$ in comparison to our Base drifting model for typical PS1 SN redshifts. \mr{This means that the SN bias-corrected standardized magnitude estimated at redshift affected by selection effect would be biased by ill-modeling of the true underlying stretch distribution.}

@@

Nonetheless, it is enlightening to study the expected mean stretch difference as
a function of redshift when considering that the underlying SN stretch
distribution follows either the per-sample Asymmetric Gaussian modeling or our Base drifting model. This
is shown Fig.~\ref{fig:magdrift}, where, for clarity we convert this difference
in mean stretch $x_1$ to a difference in magnitude using $\alpha=0.156$
\citep{scolnic2018a}. \mr{We strengthen here that this does not corresponds to the actual impact on cosmological distances derived from SNeIa depending on which underlying stretch distribution modeling as it does not account for the amplitude of the Malmquist bias correction as a function of redshift and surveys.}

@@

    \item The underlying SN~Ia stretch distribution is         redshift dependent, as previously suggested by e.g.~\cite{howell2007}. This
        result is largely independent of details on each age-population
        modeling. \yc{\textit{Indeed, and this should be stated first!}}

@@

    \item 
        \mr{Using} survey-based asymmetric Gaussian distributions, as  done, e.g., in the current implementation of the BBC, is excluded to be good descriptions of the data \nn{relative to} our drifting model. Hence, the sample-based approach does not accurately
        account for redshift drift.

@@

    \item Given the current dataset, we suggest the use of the following stretch
        population model as a function of redshift:
        \begin{align}\label{eqconclusion:stretchz}
            X_1\left(z \right) =
             &\,\delta(z)\times\mathcal{N}(\mu_1,\sigma_1^2)\,+\nonumber\\
            (1-&\,\delta(z)) \times  \left[a\times\mathcal{N}(\mu_1,\sigma_1^2) +
            (1-a)\times\mathcal{N}(\mu_2,\sigma_2^2)\right]
        \end{align}
        with: $a=0.51$, $\mu_1=0.37$, $\mu_2=-1.22$, $\sigma_1=0.61$,
        $\sigma_2=0.56$ (see Table~\ref{tab:modelresults}) and using the
        age-population drift modelization $\delta(z)$ defined in with $K=0.87$:
        \begin{align}
            \delta(z) & = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}.
        \end{align}
\end{enumerate}

@@

\begin{acknowledgements}
    This project has received funding from the European Research Council (ERC)
    under the European Union's Horizon 2020 research and innovation programme
    (grant agreement n 759194 - USNAC).
    % How has received that support for this project ? 
    %Support has been provided by the Institut Universitaire de France, the CNES, and the region Auvergne-Rhone-Alpes.
\end{acknowledgements}

@@

@@

>>>>

\author{N. Nicolas
    \thanks{n.nicolas@ipnl.in2p3.fr, equal contribution}
    \inst{1}
    \and M. Rigault
    \thanks{m.rigault@ipnl.in2p3.fr, equal contribution}
    \inst{2}\and\\
    Y. Copin\inst{1}   
    \and R. Graziani\inst{2}
    %HERE ALPHABETIC ORDER
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and J. Nordin\inst{4}
    \and Y.-L. Kim\inst{1}
    \and Saul Perlmutter\inst{3}
    % Gerard ne souhaite pas etre co-auteur.
}

\institute{Université de Lyon, F-69622, Lyon, France; Université de Lyon
    1, Villeurbanne; CNRS/IN2P3, Institut de Physique des Deux Infinis, Lyon
    \and 
    Université Clermont Auvergne, CNRS/IN2P3, Laboratoire de
    Physique de Clermont, F-63000 Clermont-Ferrand, France.
    \and
    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720 
    \and
    Institut fur Physik, Humboldt-Universität zu Berlin, Newtonstr. 15, 12489 Berlin
}

@@

In this paper, we take a step aside to probe the validity of our modeling of the
SN population, which we claim to be constituted of two age-populations
\citep{rigault2013,rigault2015,rigault2018}: one old and one younger, the former
having on average lower lightcurve stretches and being brighter after
standardization. We use the correlation between the SN age, as probed by the LsSFR, and the SN stretch to model
the expected evolution of the underlying SN stretch distribution as a function
of redshift. This modeling relies on three assumptions: (1) there are two
distinct populations of SNe~Ia; (2) the relative fraction of each of these
populations as a function of redshift follows the model presented in
\cite{rigault2018} and (3) the underlying distribution of stretch for each age
sample is constant. This paper aims at testing this specific model with datasets from the literature. \mr{Note that the progenitor age as traced by the LsSFR seems to capture physical feature intrinsic to the progenitor and/or explosion mechanism that the stretch alone is not capturing \citep{nordin2018}.}

@@

SNLS typically acquires SNe~Ia in the redshift range $0.4<z<0.8$; at these
redshifts, the rest-frame Bessel $B$ band roughly corresponds to the SNLS $i$
filter, that has a $5\sigma$ depth of
24.8~mag\footnote{\href{https://www.cfht.hawaii.edu/Science/CFHTLS/cfhtlsfinalreleaseexecsummary.html}{CFHT
final release website.}}. This converts to a $z_{lim}=0.60$, in agreement with
\cite{neill2006}, \cite{perrett2010} and \cite{bazin2011}.  Fig.~14 of
\cite[][\mr{see their section~5}]{perrett2010} suggests a lower limit of $z_{lim}=0.55$; both limits will be
considered, as discussed below.

@@

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008,sako2008}, which would lead to a $z_{lim}=0.24$. However, the
SDSS surveys were more sensitive to limited spectroscopic resources;
\cite[][\mr{see their section~2}]{kessler2009} pointed out that during the first year of SDSS, SNe~Ia with
$r-mag<20.5$ were favored for spectroscopic follow-up, corresponding to a
redshift cut at $0.15$. For the rest of the SDSS survey, additional
spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a \mr{reasonable} completeness up to
$z_{lim}=0.2$. Following these analyses, we will use $z_{lim}=0.2$ as the
baseline SDSS redshift limit.

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift distribution of these three surveys is shown in Fig.~\ref{fig:cuts}.
As expected, the selected redshift limits roughly correspond to the peak of
these histograms. \mr{We show in section~\ref{sec:results} that this sample selection indeed provide a subset of SNe~Ia with insignificant selection effects when compared to state-of-the-art Malmquist correction techniques.}

@@

We also performed tests allowing
the high-stretch mode of the old population to differ from the young population mode, hence adding two degrees of freedom. The corresponding fit is not
significantly better, with a $\Delta$AIC of $-0.4$. 
this strengthens the fact
that the young and old populations indeed appear to share the same underlying
high-stretch mode. 
\mr{Furthermore, one might wonder whether a low-stretch mode might also exist in the young-population, see Fig.~\ref{fig:stretchlssfr}. We tested that by allowing this population to also be bi-modal, finding the amplitude of this low-stretch mode to be compatible with~0 in this young population ($<2\%$). More generally, this raises the question of inaccurate tracing of age by a given environmental tracer (here the LsSFR). A dedicated analysis be presented in Briday et al. in prep.}

@@

Commonly used Malmquist bias correction techniques, such as the BBC-formalism, assume per Sample Asymmetric Gaussian functions for modeling the underlying stretch and color distributions.
Yet, as shown in Section~\ref{sec:results}, such a sample-based distribution is excluded as being as good as our drifting model.
Then, unlike what \citet[][section~2]{scolnic2016} and
\citet[][section~5.4]{scolnic2018a} suggested, i.e. that
traditional surveys span limited redshift ranges and that therefore the per-sample approach accounts for implicit redshift drifts, \sout{might not be correct} \nn{a direct modeling of the redshift drift is more appropriate than a sample-based approach}. We stress here that, as measurements of modern surveys try to cover increasingly larger redshift ranges in order to reduce calibration systematic uncertainties, this sample-based approach is becoming less valid, notably for PS1, DES and, soon, LSST.

We illustrate in Fig.~\ref{fig:bbc_pdf_ps1} the prediction difference in the
underlying stretch distribution between the per-sample asymmetric modeling and our Base drifting
model for the PS1 sample. Our model is bimodal and the relative amplitude of
each mode depends on the redshift-dependent fraction of old and young SNe~Ia in
the sample: the higher the fraction of old SNe~Ia (at lower redshift), the
higher the amplitude of the old-specific low-stretch mode. This redshift
dependency is shown as blue to red underlying distributions in
Fig.~\ref{fig:bbc_pdf_ps1} for redshift ranges covered by PS1. The observed
$x_1$ histogram follows our modeling defined as the sum of individual underlying SN-redshift distributions. As expected, the two modeling approaches differ mostly in the negative part of the SN stretch distribution. The asymmetric Gaussian distribution goes through the middle of the
bimodal distribution, over-estimating the number of SNe~Ia at $x_1\sim-0.7$ and
under-estimating it at $x_1\sim-1.7$ in comparison to our Base drifting model for typical PS1 SN redshifts. \mr{This means that the SN bias-corrected standardized magnitude estimated at redshift affected by selection effect would be biased by an ill-modeling of the true underlying stretch distribution.}

@@

Nonetheless, it is enlightening to study the expected mean stretch difference as
a function of redshift when considering that the underlying SN stretch
distribution follows either the per-sample Asymmetric Gaussian modeling or our Base drifting model. This
is shown Fig.~\ref{fig:magdrift}, where, for clarity we convert this difference
in mean stretch $x_1$ to a difference in magnitude using $\alpha=0.156$
\citep{scolnic2018a}. \mr{We stress here that this does not correspond to the actual impact on cosmological distances derived from SNeIa as this simple stretch average comparison does not account for the amplitude of the Malmquist bias correction to be applied as a function of redshift and surveys.}

@@

    \item The underlying SN~Ia stretch distribution is         redshift dependent, as previously suggested by e.g.~\cite{howell2007}. This
        result is largely independent of details on each age-population
        modeling.

@@

    \item 
        \mr{Using} survey-based asymmetric Gaussian distributions, as  done, e.g., in the current implementation of the BBC, is excluded to be good descriptions of the data \nn{relative to} our drifting model. Hence, the sample-based approach does not accurately account for redshift drift and even less so as survey span larger redshift ranges. \mr{We warn that, even if extra degrees of freedom might be acceptable SN cosmological studies, extrapolating the SN property distributions from the volume-limited part of a survey to the magnitude-limited one affected by sample selection would still be inaccurate since the distribution would drift.}

@@

    \item Given the current dataset, we suggest the use of the following stretch
        population model as a function of redshift:
        \begin{align}\label{eqconclusion:stretchz}
            X_1\left(z \right) =
             &\,\delta(z)\times\mathcal{N}(\mu_1,\sigma_1^2)\,+\nonumber\\
            (1-&\,\delta(z)) \times  \left[a\times\mathcal{N}(\mu_1,\sigma_1^2) +
            (1-a)\times\mathcal{N}(\mu_2,\sigma_2^2)\right]
        \end{align}
        with: $a=0.51$, $\mu_1=0.37$, $\mu_2=-1.22$, $\sigma_1=0.61$,
        $\sigma_2=0.56$ (see Table~\ref{tab:modelresults}) and using the
        age-population drift modelization $\delta(z)$ with $K=0.87$:
        \begin{align}
            \delta(z) & = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}.
        \end{align}
\end{enumerate}

@@

\begin{acknowledgements}
    This project has received funding from the European Research Council (ERC)
    under the European Union's Horizon 2020 research and innovation programme
    (grant agreement n 759194 - USNAC).
    This work was supported in part by the Director, Office of Science, Office of High Energy Physics of the U.S. Department of Energy under Contract No. DE-AC025CH11231.
    % How has received that support for this project ? 
    %Support has been provided by the Institut Universitaire de France, the CNES, and the region Auvergne-Rhone-Alpes.
\end{acknowledgements}

@@

\bibitem[Nordin et al.(2018)]{nordin2018} Nordin, J., Aldering, G., Antilogus, P., et al.\ 2018, \aap, 614, A71

@@

\bibitem[Saunders et al.(2020)]{saunders2020} Saunders, C., Aldering, G., Antilogus, P., et al.\ 2020, VizieR Online Data Catalog, J/ApJ/869/167
